---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
First we add the required libraries.
```{r install libraries, echo = TRUE,  message=FALSE}
library(ggplot2)
library(tidyverse)
library(lubridate)
```

## Loading and preprocessing the data
Then we load the data into a table we can work with.
```{r unzip and read data, echo = TRUE}
if (!file.exists("activity.csv"))
    unzip("activity.zip")
activity <- read_csv("activity.csv")
activity$date <- ymd(activity$date)
```


## What is mean total number of steps taken per day?
Then we summarize by day and take the average

```{r make step histogram, echo = TRUE}
daystep <- activity %>% 
  group_by(date) %>% 
  summarize(steptotal = sum(steps))
ggplot(daystep, aes(x=date, y = steptotal)) +
  geom_histogram(stat="identity")
```
<br/>Then we calculate the mean over all days.

```{r calculate mean and median daily steps, echo = TRUE}
meanstep <- as.character(signif(mean(daystep$steptotal, na.rm=TRUE)), digits = 6)
paste("The mean steps per day was",meanstep)
medianstep <- as.character(median(daystep$steptotal, na.rm = TRUE))
paste("The median steps per day was",medianstep)
```


## What is the average daily activity pattern?
We cut out the date and then group by interval to make an average number of steps for each interval.

```{r calculate daily activity pattern, echo = TRUE}
dailyact <- activity %>%
  select(-c(date)) %>%
  group_by(interval) %>%
  summarize(averagestep = mean(steps, na.rm = TRUE)) 
ggplot(dailyact, aes(x=interval, y=averagestep))  +geom_line()
```

```{r calculate highest average interval, echo = TRUE}
mxavg <- dailyact[which.max(dailyact$averagestep),]
mxintname <- as.character(mxavg[1])
paste("The interval that has the highest amount of average steps is", mxintname)
```
  
## Imputing missing values
First step is to impute the missing values by calculating the average step per
interval as a column variable, then we just assign that average to the missing value. It's not exactly a good statistical solution but it does the job here.

```{r calculate missing values, echo = TRUE}
naamt <- (sum(is.na(activity$steps)))
paste("The number of NA values is", naamt)
gactivity <- activity %>% group_by(interval) %>% mutate(step_avg = round(mean(steps,na.rm=T)))
gactivity$steps[is.na(gactivity$steps)] <- gactivity$step_avg[is.na(gactivity$steps)]
```

Then we use our new calculation to make a histogram.

``` {r calculate new histogram with imputed values, echo = TRUE}
ndaystep <- gactivity %>% 
  group_by(date) %>% 
  select(-c("step_avg")) %>%
  summarize(steptotal = sum(steps))
ggplot(ndaystep, aes(x=date, y = steptotal)) +
  geom_histogram(stat="identity")
```

Finally we report back the new median and mean. This shouldn't be different
because we were using the mean itself to fill in the missing values, so it doesn't create any variance.

```{r print new median and mean step, echo = TRUE}
meanstep2 <- as.character(signif(mean(ndaystep$steptotal, na.rm=TRUE)), digits = 6)
paste("The new mean steps per day was",meanstep2)
medianstep2 <- as.character(median(ndaystep$steptotal, na.rm = TRUE))
paste("The new median steps per day was",medianstep2)
```

## Are there differences in activity patterns between weekdays and weekends?
We create a vector of weekdays and test them against the weekday value stored
in the date variable, and then factor based on its existance. Then we plot it.

```{r create weekday vs weekend plot, echo = TRUE}
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')

gactivity$isweekday <- factor((weekdays(gactivity$date) %in% weekdays1), 
    levels = c(FALSE, TRUE), labels=c('weekend','weekday')) 

gactivity <- select(gactivity, -c("step_avg"))

ggplot(gactivity, aes(x= interval)) + 
  geom_line(color = "blue",alpha=0.66, aes(y=steps), size=.5) + 
  facet_wrap(~ isweekday, nrow=2, ncol=1) +
  labs(x = "time interval", y = "Number of Steps")
```
 
